---
title: "ProbabilityAssessment"
date: "`r Sys.Date()`"
output: html_document
---
# load packages
```{r}
library(ggplot2)
library(patchwork)
library(dplyr)
library(readxl)
library(tidyr)
```

# Assessment 1: Statistical hypothesis testing
# Question: do the levels at inclusion vary between males and females?

## prepare the table
```{r}
# read the table biomarkers and covariates tables from local
biomarker <- read_excel("biomarkers.xlsx")
covariates <- read_excel("covariates.xlsx")

# filter the biomarker table with time point = 0weeks
biomarker <- biomarker %>%
  separate(Biomarker, into = c("PatientID", "TimePoint"), sep = "-", remove = T) %>% 
  filter(TimePoint=="0weeks")

Sex <- covariates[,c(1,3)]

# check if the patients in two tables match each other
missing_patient <- setdiff(Sex$PatientID, biomarker$PatientID)
print(missing_patient)

# remove patient with ID 40 which missing biomarker information
Sex <- Sex[!(Sex$PatientID %in% missing_patient), ]

# add age to the biomarker table
merged_data <- merge(biomarker, Sex, by = "PatientID")
names(merged_data)[names(merged_data) == 'Sex (1=male, 2=female)'] <- 'Sex'
merged_data$Sex <- recode(merged_data$Sex,
                          `1` = "male",
                          `2` = "female")

merged_data$Sex <- factor(merged_data$Sex, levels = c("male", "female"))
biomarkersName <- c("IL-8","VEGF-A","OPG","TGF-beta-1","IL-6","CXCL9","CXCL1","IL-18","CSF-1")

```

## Assess Normality
```{r}
# write a function for testing normality
check_normality <- function(data, biomarkersName) {
  male_data <- data %>% filter(Sex == "male") %>% pull(biomarkersName)
  female_data <- data %>% filter(Sex == "female") %>% pull(biomarkersName)
  male_shapiro <- shapiro.test(male_data)
  female_shapiro <- shapiro.test(female_data)
  return(data.frame(
    Biomarker = biomarkersName,
    Male_p = male_shapiro$p.value,
    Female_p = female_shapiro$p.value
  ))
}

# apply the function to all biomarkers
normality_results <- lapply(biomarkersName, function(b) check_normality(merged_data, b))
normality_resultTable <- do.call(rbind, normality_results)
print(normality_resultTable)

# Add columns indicating normality to the table based on p-values
normality_resultTable <- normality_resultTable %>%
  mutate(
    Male_normal = Male_p > 0.05,
    Female_normal = Female_p > 0.05
  )

# write.csv(normality_resultTable,file = "normality_result.csv", row.names = T)  
```

## perform the t.test or Wilcoxon test
```{r}
# write a function for the hypothesis test
perform_test <- function(data, biomarker, normality_info) {
  male_data <- data %>% filter(Sex == "male") %>% pull(biomarker)
  female_data <- data %>% filter(Sex == "female") %>% pull(biomarker)
  
  # Retrieve normality results for the biomarker
  normality_row <- normality_info %>% filter(Biomarker == biomarker)
  
  male_normal <- normality_row$Male_normal
  female_normal <- normality_row$Female_normal
  
  # Decide on the test based on normality
  if (male_normal & female_normal) {
    # Check equality of variances
    var_test <- var.test(male_data, female_data)
    var_equal <- var_test$p.value > 0.05
    # Perform t-test
    t_result <- t.test(male_data, female_data, var.equal = var_equal)
    test_used <- ifelse(var_equal, "t-test (equal variances)", "Welch's t-test (unequal variances)")
    p_value <- t_result$p.value
  } else {
    # Perform Wilcoxon rank-sum Test
    w_result <- wilcox.test(male_data, female_data)
    test_used <- "Wilcoxon"
    p_value <- w_result$p.value
  }
  # Return the results
  return(data.frame(
    Biomarker = biomarker,
    Test = test_used,
    p_value = p_value
  ))
}

# Step 3: Apply the function to all biomarkers

test_results_list <- lapply(biomarkersName, function(b) {
  perform_test(merged_data, b, normality_resultTable)
})

test_results_table <- do.call(rbind, test_results_list)
test_results_table

# write.csv(test_results_table,file = "test_result1.csv", row.names = T)
```

## Type 1 error rate with Multiple Testing
```{r}
# total number of biomarkers
n <- 9
# Significance level for each individual test
alpha <- 0.05
# the probability of not making a Type I error in a single test
p_NoError_single <- 1 - alpha
print(p_NoError_single)
# probability of making at least one Type I error
p_atLeastOneError <- 1 - p_NoError_single^n
print(p_atLeastOneError)
```

## Adjust for multiple testing by Bonferroni Correlation
```{r}
test_results_tableAdj <- test_results_table
test_results_tableAdj$Bonferroni_p <- p.adjust(test_results_tableAdj$p_value, method = "bonferroni")
test_results_tableAdj$Significant_Bonferroni <- test_results_tableAdj$Bonferroni_p < 0.05
test_results_tableAdj

write.csv(test_results_tableAdj,file = "test_resultAdj.csv", row.names = T)

```

# Assessment 2: Regression modelling
## read the table biomarkers and covariates tables from local
```{r}
biomarker <- read_excel("biomarkers.xlsx")
covariates <- read_excel("covariates.xlsx")

# filter the biomarker table with time point = 0weeks
biomarker <- biomarker %>%
  separate(Biomarker, into = c("PatientID", "TimePoint"), sep = "-", remove = T) %>% 
  filter(TimePoint=="0weeks")

# check if the patients in two tables match each other
missing_patient <- setdiff(covariates$PatientID, biomarker$PatientID)
print(missing_patient)

# remove patient with ID 40 which missing biomarker information
covariates <- covariates[!(covariates$PatientID %in% missing_patient), ]
```

## merge two tables, and then modify and filter the combined_data
```{r}
combined_data <- merge(covariates, biomarker, by = "PatientID")
# rename the rownames
colnames(combined_data) <- gsub("-", "_", colnames(combined_data))
colnames(combined_data)[colnames(combined_data) == "Sex (1=male, 2=female)"] <- "Sex"
colnames(combined_data)[colnames(combined_data) == "Smoker (1=yes, 2=no)"] <- "Smoker"
# remove the rows with Vas_12months == NA
combined_data <- combined_data %>%
  filter(!is.na(Vas_12months))
head(combined_data)
```

## randomly subset 80% patients from the combined_data for building the regression model, the rest for verification
```{r}
set.seed(222) 
shuffled_data <- combined_data[sample(nrow(combined_data)), ]
train_size <- floor(0.8 * nrow(shuffled_data))
lm_data <- shuffled_data[1:train_size, ]
test_data <- shuffled_data[(train_size + 1):nrow(shuffled_data), ]
# Check the sizes of two tables
cat("Number for lm:", nrow(lm_data), "\n")
cat("Number for test:", nrow(test_data), "\n")
```

## Fit the regression model
```{r}
model <- lm(Vas_12months ~ IL_8 + VEGF_A + OPG + TGF_beta_1 + IL_6 + CXCL9 + CXCL1 +IL_18 + CSF_1 + Age + Sex + Smoker, data = lm_data)
summary(model)
# save the table
model_summary <- summary(model)
coefficients <- model_summary$coefficients
coef_table <- as.data.frame(coefficients)
colnames(coef_table) <- c("Coefficient Estimate", "Standard Error", "t-value", "p-value")
# View the table
print(coef_table)
write.csv(coef_table,file = "Task2coef_table.csv", row.names = T)
```

## calculate the residuals, mean squared rrror and R-squared of the model
```{r}
lm_predictions <- predict(model, newdata = lm_data)
lm_residuals <- lm_data$Vas_12months - lm_predictions
# Mean Squared Error and R-squared 
mean(lm_residuals^2)  
summary(model)$r.squared  

```

## Predictions on the test_data
```{r}
test_predictions <- predict(model, newdata = test_data)
# predicted vs actual 12-month VAS in the test set
comparison <- data.frame(Actual = test_data$Vas_12months, Predicted = test_predictions)
print(comparison)
write.csv(comparison,file = "Task2comparison.csv", row.names = T)
# Plot actual vs predicted values
p1 <- ggplot(comparison, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  ggtitle("Actual vs. Predicted 12-month VAS on test data")
p1
# Mean Squared Error and R-squard
mean((test_data$Vas_12months - test_predictions)^2)
cor(test_data$Vas_12months, test_predictions)^2  

```

